---
title: "Notes"
author: "Sami Rifai"
date: "6/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)
```

# Non-linear model forms  
 Asym-Drop*exp(-exp(lrc)*mape^pwr) + 
 B1*(pe_anom_12mo/mape) + 
 B2*(cco2*mape)+ 
 B3*(cco2*pe_anom_12mo/mape) + 
 B4*as.numeric(epoch)


# Weibull form
$$NDVI_{3 mo}=Asym-Drop[exp(-exp(lrc)\,(\frac{P}{PET_{MA}})^{pwr}]+\eta$$
# eta linear modifier
$$\eta =  \frac{B_1\,\frac{P}{PET_{anom}}}{\frac{P}{PET_{MA}}} + 
 \frac{B_2\,CO_{2}\,P}{PET_{MA}}+ 
 \frac{B_3\,CO_{2}\,\frac{P}{PET_{anom}}}{\frac{P}{PET_{MA}}} + 
 B4\,sensor$$


Sharpe-Schoolfield equation: 
$log(rate) = lnc + E(\frac{1}{T_{c}} - \frac{1}{kT}) - ln(1 + e^{E_h(\frac{1}{kT_h} - \frac{1}{kT})})$

# Asymptotic function 
$$NDVI = A + (B[0]-A)*exp(-exp(\beta)*\frac{P}{PET})$$

# Logistic function
$$NDVI =  \frac{Asym}{(1+exp(\frac{(xmid-\frac{P}{PET})}{scal}))}$$

# Richards growth function
$$NDVI = \frac{A_1*(1+exp(B_1 - P:PET)}{C^{-exp(-D_1)}}$$

# Modified Richards equation
ric_x2 <- function(x,x2,
                   Asym,Asym2,
                   xmid,xmid2,
                   scal,scal2,
                   lpow,lpow2){
  (Asym+Asym2*x2) * (1+exp(((xmid+xmid2*x2) - x)/(scal+scal2*x2)))^(-exp(-(lpow+lpow2*x2)))
}

$(A_1+A_2*CO_2)*(1+exp(((B_1+B_2*x2) - P:PET)/(C_1+C_2*x2)))^(-exp(-(D_1+D_2*x2)))$

$$NDVI = \frac{(A_1+A_2*CO_2)*(1+exp(((B_1+B_2*CO_2) - P:PET)}{(C_1+C_2*CO_2)))^(-exp(-(D_1+D_2*CO_2)))}$$

$NDVI_{MCD64} = s(sz) + s(NIR) + s(RED) + s(NDVI_{AVHRR}) + s(x,y)$


$$\frac{A}{E} = \frac{C_{a}(1-\frac{C_i}{C_a})}{1.6VPD}$$


<!-- m4_ndvi <- bam(ndvi_mcd~ -->
<!--                  s(sz)+s(nir_c)+s(red_c)+s(ndvi_c,m = 1)+ -->
<!--                  s(x,y)+s(vc,bs='re'), -->
<!--                family=betar(link='logit'), -->
<!--                select = TRUE, discrete=TRUE, method='fREML', nthreads = 8, -->
<!--                data=train) -->






From Donohue et al 2013 
$$W_p = \frac{A_l}{E_l} = \frac{C_a}{1.6D}(1 - \frac{C_i}{C_a})$$ 


\begin{equation}
  \tag{1}
W_p = \frac{A_l}{E_l} = \frac{C_a}{1.6D}(1 - \frac{C_i}{C_a})
\end{equation}


\begin{equation}
  \tag{2}
  $$\frac{dW_p}{W_p}=\frac{dA_l}{A_l} - \frac{dE_l}{E_l} = \frac{dC_a}{C_a} - \frac{dD}{D} + \frac{d(1-\chi)}{(1-\chi)}$$
\end{equation}

\begin{equation}
$$\chi \approx \sqrt(D)$$
\end
\begin{equation}
  \tag{3}
    $$\frac{dW_p}{W_p}=\frac{dA_l}{A_l} - \frac{dE_l}{E_l} = \frac{dC_a}{C_a} - \frac{1}{2}\frac{dD}{D}$$
\end{equation}

\begin{equation}
  \tag{5}
  $$-\frac{dE_l}{E_l} = \frac{dL}{L}$$
\end{equation}

\begin{equation}
  \tag{6}
      $$\frac{dW}{W} \approx  \frac{dA_l}{A_l} + \frac{dF}{F} \approx \frac{dC_a}{C_a} - \frac{1}{2}\frac{dD}{D}$$
\end{equation}

\begin{equation}
  \tag{}
\end{equation}

\begin{equation}
  \tag{}
\end{equation}



$$W = \frac{A}{E} = \frac{C_a}{1.6D}(1 - \chi)$$ 


# NDVI year model modifier
$$NDVI = \beta_0 + \beta_1\,year + 
\beta_2\,Precip_{anom}+
\beta_3\,P:PET_{anom}+
\beta_4\,sensor$$ 
 
100*(38*b1)/b0
$$100*[\frac{\beta_1(2019-1982)}{\beta_0}]$$
 
$$\Delta NDVI \approx \frac{1}{2}[\frac{dCa}{Ca}-\frac{dD}{2\,D}]$$
 
 
 
$$NDVI=\beta_0+ \beta_1\,year+\beta_2\,P_{anom}+\beta_3\,(\frac{P}{PET})_{anom}+\beta_4\,sensor$$ 

ndvi_3mo ~ 
                                 Asym-Drop*exp(-exp(lrc)*mape^pwr) + 
                                 B1*(pe_anom_12mo/mape) + 
                                 B2*(cco2*mape)+ 
                                 B3*(cco2*pe_anom_12mo/mape) + 
                                 B4*as.numeric(epoch)

$$\eta =  \frac{B_1\,(\frac{P}{PET})_{anom}}{(\frac{P}{PET})_{MA}} + B_2\,CO_{2}\,(\frac{P}{PET})_{MA}+\frac{B_3\,CO_{2}\,(\frac{P}{PET})_{anom}}{(\frac{P}{PET})_{MA}}+B_{4}\,sensor$$
$$\frac{B_3\,CO_{2}\,(\frac{P}{PET})_{anom}}{(\frac{P}{PET})_{MA}}$$





















